\documentclass[12pt, a4paper]{article}
\input{setup}

\date{Seminar date}
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TITLE PAGE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{title}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\begin{abstract}
\setstretch{1}
\noindent 
While the body of literature on the non-take-up of public aid has grown substantially in recent years, a notable gap remains in the literature of non-take-up rates for student aid programs, where research is still extremely limited.
This paper examines the non-take-up rate of Germany’s federal student aid program BAföG by creating a microsimulation based on data from the German Socio-Economic Panel for the period 2007--2021.
Using the outcome of our microsimulation, we estimate three specifications of binary choice models to examine how individual characteristics relate to take-up decisions. 
Our findings indicate that non-take-up has increased over the past decade, with an average rate of approximately 60\% for our study period. 
Several factors contribute to this pattern. Students who expect only a small award are much more likely to forgo BAföG, while those who are more familiar with the programme, for example through a sibling who has already claimed, tend to have lower non-take-up rates. We also observe notable regional differences as students from East Germany are more likely to apply than those from West Germany, which may reflect differing attitudes toward state support. Age and partnership status are also associated with higher non-take-up, whereas we do not find evidence that behavioural traits such as risk preference, patience, or impulsiveness play a substantial role.
\vspace{1cm} \\
    \noindent \textbf{Keywords:} Non-take-up, 
    Microsimulation, 
    SOEP, 
    Student Aid, 
    Student Loans, 
    BAföG, 
    Education Funding. \\
    \noindent \textbf{JEL codes:} I22, I23, I24, I38, H53
\end{abstract}

\newpage


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ACKNOWLEDGEMENTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{center}
    \textbf{Acknowledgements}
\end{center}
We would like to thank our supervisor, Petra Thiemann, for her helpful guidance and support throughout the writing of this paper. 
Her feedback was valuable in shaping the direction and clarity of our work.

\newpage
\begingroup
    \setstretch{1.1}
    \hypersetup{hidelinks}
    \tableofcontents
\endgroup
\thispagestyle{empty}

\newpage
\setcounter{page}{1}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CONTENT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{sections/1 Intro}
\input{sections/2 Literature}
\input{sections/3 Background}
\input{sections/4 Data}
\input{sections/5 Method}
\input{sections/6 Results}
\input{sections/7 Discussion}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% REFERENCES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{flushleft}
    \renewcommand{\baselinestretch}{1.1}
    \small
    \titleformat{\section}{\normalfont\huge\bfseries}{\thesection}{1em}{} 
    \addcontentsline{toc}{section}{References} 
    \bibliography{bibliography, bafogandg, estg}
\end{flushleft}

%APPENDICES
\newpage
\appendix
\setcounter{page}{1}        % Restarts page counting to 1
\pagenumbering{roman}       % Differentiates the appendix page numbers from the main article

\titleformat{\section}{\centering\normalfont\normalsize\bfseries}{Appendix \thesection: }{0em}{}
\input{appendix/appendix_tables}
\input{appendix/appendix_figures}
\input{appendix/appendix_variable_dictionary}
\input{appendix/appendix_example_simulation}
\input{appendix/appendix_input_params}

\newpage
\pagenumbering{arabic}                          % Return to standard page numbering (optional)
\renewcommand{\thesection}{\arabic{section}}    % Reset section numbering format if needed
\setcounter{section}{0}                         % Reset section counter if using \section in ai_statement


\input{ai_statement}
\end{document}
